Experiences with Planning for Natural Language Generation

### 2008-06-23 trunk/Papers/experiences/domains.tex
I'm essentially happy now

In practice, the process of determining referring expressions typically
interacts with the realization step and so it turns out to be beneficial to
perform both of these steps together. This was the goal of the SPUD system
\cite{Stone2003a}, which performed both steps together using a top-down
generation algorithm based on tree-adjoining grammars \cite{joshi;etal1997}
whose lexical entries were equipped with semantic and pragmatic
information. Unfortunately, SPUD suffered from having to explore a huge
search space, and had to resort to a non-optimal greedy search strategy to
retain reasonable efficiency.

%However, determining referring expressions and realization
%interact, so it would be beneficial to perform both steps together.
%This was the goal of the SPUD system \cite{Stone2003a}, which
%performed both steps together using a top-down generation algorithm
%based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
%entries were equipped with semantic and pragmatic information.
%Unfortunately, SPUD suffered from having to explore a huge search
%space, and had to resort to a non-optimal greedy search strategy to
%retain reasonable efficiency.

@@

%\citeauthor{KolSto07}~\shortcite{KolSto07} attempt to improve the
%efficiency by translating the SPUD sentence generation problem into a
%planning problem and using a planning algorithm for generation.  We will
%now illustrate how they do this by a (simplified) example.

To improve on the efficiency of SPUD,
\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
generation problem into a planning problem and use a planning algorithm for
generation. We illustrate this process on a simplified example.

Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
shown in Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can instantiate
these trees by substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
shown in Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
derivation by starting with the elementary tree for ``sleeps''. This tree
satisfies the need to convey the semantic information, but introduces a
need to generate a noun phrase (NP) for the subject, which also refers
uniquely to the target referent $r_1$. In a second step, SPUD substitutes
the tree for ``the rabbit'' into the open NP leaf, which makes the
derivation grammatically complete. Since there are two different
individuals that could be described as ``the rabbit''---technically, $r_2$
is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
the referring expression---we are still not finished. To complete the
derivation, the tree for ``white'' is added to the existing structure by
the \emph{adjunction} operation, making the derivation syntactically and
semantically complete.

@@

A map of an example GIVE world is shown in
Figure~\ref{fig:give-development-world}.  In this world, the user's task is
to pick up a trophy in the top left room.  The trophy is hidden in a safe
behind a picture, so the user must first move the picture out of the way
and open the safe by pushing a sequence of buttons (the small square boxes
on the walls) in the correct order.  In order to get to all these buttons,
the user must also open the lower door and deactivate the alarm tile by
pushing further buttons.  To simplify both the planning and the NLG task,
the world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next. Figure~\ref{fig:give-planning} shows some of the
available actions in PDDL syntax.

@@

The plans generated in the GIVE domain are often nontrivial. For instance,
in the example above the shortest plan consists of 108 action steps; the
first few steps are as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}
%
(Most of the actions in this plan are ``move'' actions.) The bulk of the
GIVE problem can be considered a variant of the Gridworld problem, which
also involves finding a route through a world with discrete positions---but
with the additional need to press buttons in the right order, reason about
many more objects in the world, and navigate somewhat more complicated room
shapes.

The task of generating natural language instructions in this domain is also
more complicated than just a straightforward translation of the action
steps in a plan. For instance, the NLG system could generate an instruction
sequence starting with ``turn left; walk forward; press the button; turn
right; walk forward; walk forward; turn right; walk forward; walk forward;
turn left; walk forward''.  However, this would be clumsy and boring, and
it is to be expected that frustrated users will quickly cancel an
interaction with such a system, resulting in a bad evaluation score.
Having just pressed the button, it would be much better to say ``turn
around and walk through the door''.  This means that the mapping from
action instances to natural language instructions is not trivial.
Conversely, it may also be necessary to express a single planning step with
several instructions.  Having just entered the top left room, it may be
easier for the user to understand the instruction ``walk to the centre of
the room; turn right; now press the green button in front of you'' than the
instruction ``press the green button on the wall to your right''.  To plan
the referring expression ``the green button in front of you'' at a time
when the button is not yet in front of the user, the NLG system must keep
track of the hypothetical changes in what objects are visible to the user.
Thus the acts of referring and instructing, and the modules for planning
and NLG, must be tightly integrated.

Plan execution monitoring also plays an important role in the GIVE problem.
At a high level, the system needs to monitor a user's actions and compare
them against the generated instruction set to determine if the user has
correctly followed directions or not. In the case of the latter, new
instructions may have to be generated. In practice, the situation
can be quite complicated since the mental state of the user is not known
and so the system must observe the user's actions in real time. For
example, a user directed to ``turn around and walk through the door'' may
not necessarily perform these actions to the letter, i.e., immediately
turning 180 degrees and proceeding directly to the door. Instead, the user
might take a roundabout route through the room, eventually exiting out the
door. Although the user's actions do not match the generated instructions
exactly, they meet the intended goal. The system must be able to identify
such ``equivalent'' plans and not immediately generate new instructions as
soon as the user's actions have gone off course. Furthermore, a user can
communicate certain intentions to the system, both through action and
inaction. For instance, the system should infer that a user has failed to
follow instructions if the user exits a room when given a directive to
``walk to the centre of the room''. However, the system should also make a
similar conclusion if a user simply does nothing when given the
instruction. Such real-time plan execution monitoring is essential in this
domain.

%However, the system should also be able to make a similar conclusion if a
%user simply does nothing when given the instruction. Thus, such real-time
%plan monitoring is essential in the GIVE domain, and also affects other
%modules in the system.

>>>>

In practice, the process of determining referring expressions
typically interacts with the realization step and so it turns out to
be beneficial to perform both of these steps together. This was the
goal of the SPUD system \cite{Stone2003a}, which did this using a
top-down generation algorithm based on tree-adjoining grammars
\cite{joshi;etal1997} whose lexical entries were equipped with
semantic and pragmatic information. Unfortunately, SPUD suffered from
having to explore a huge search space, and had to resort to a
non-optimal greedy search strategy to retain reasonable efficiency.
To improve on the efficiency of SPUD,
\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
generation problem into a planning problem and use a planning
algorithm for generation.

@@

We illustrate this process on a simplified example.  Consider a
knowledge base containing the individuals $e$, $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits,
$r_1$ is white and $r_2$ is brown, and $e$ is an event in which $r_1$
sleeps.  Say that we want to express the information
$\{\mathsf{sleep}(e,r_1)\}$ using the tree-adjoining grammar shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can
instantiate these trees by substituting individuals for \emph{semantic
  roles}, such as $\mathsf{self}$ and $\mathsf{subj}$, and then
combine the tree instances as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''.

We compute this grammatical derivation top-down by starting with the
elementary tree for ``sleeps''. This tree satisfies the need to convey
the semantic information, but introduces a need to generate a noun
phrase (NP) for the subject; this NP must refer uniquely to the target
referent $r_1$. In a second step, we substitute the tree for ``the
rabbit'' into the open NP leaf, which makes the derivation
grammatically complete. Since there are two different individuals that
could be described as ``the rabbit''---technically, $r_2$ is still a
\emph{distractor} (i.e., based on the description ``the rabbit'', the
hearer might erroneously think that we're talking about $r_2$ and not
$r_1$)---we are still not finished. To complete the derivation, the
tree for ``white'' is added to the existing structure by the
\emph{adjunction} operation, making the derivation syntactically and
semantically complete.

@@

A map of an example GIVE world is shown in
Figure~\ref{fig:give-development-world}.  In this world, the user's
task is to pick up a trophy in the top left room.  The trophy is
hidden in a safe behind a picture, so the user must first move the
picture out of the way and open the safe by pushing a sequence of
buttons (the small square boxes on the walls) in the correct order.
In order to get to all these buttons, the user must also open the door
in the centre of the map and deactivate the alarm tile by pushing
further buttons.  To simplify both the planning and the NLG task, the
world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one
tile to the centre of the next. Figure~\ref{fig:give-planning} shows
some of the available actions in PDDL syntax.

@@

The plans generated in the GIVE domain are often nontrivial. For instance,
in the example above the shortest plan consists of 108 action steps; the
first few steps are as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}

Most of the actions in this plan are ``move'' actions.  This makes the
bulk of the GIVE problem very similar to the Gridworld problem, which
also involves finding a route through a world with discrete
positions---but with the additional need to press buttons in the right
order, reason about many more objects in the world, and navigate
somewhat more complicated room shapes.

The GIVE task comes with a number of interesting research challenges
both for the NLG and the planning communities.  On the NLG side, one
major problem is that the natural language instruction sequences can't
simply be obtained by verbalising each action instance in the plan in
turn.  For instance, the NLG system could generate an instruction
sequence starting with ``turn left; walk forward; press the button;
turn right; walk forward; walk forward; turn right; walk forward; walk
forward; turn left; walk forward''.  However, this would be clumsy and
boring, and it is to be expected that frustrated users will quickly
cancel an interaction with such a system, resulting in a bad
evaluation score.  Having just pressed the button, it would be much
better to say ``turn around and walk through the door''; i.e.,
multiple action instances should be merged into a single
instruction. Conversely, it may also be necessary to express a single
planning step with several instructions.  Having just entered the top
left room, it may be easier for the user to understand the instruction
``walk to the centre of the room; turn right; now press the green
button in front of you'' than the instruction ``press the green button
on the wall to your right''.  To plan the referring expression ``the
green button in front of you'' at a time when the button is not yet in
front of the user, the NLG system must keep track of the hypothetical
changes in what objects are visible to the user.  Thus the acts of
referring and instructing, and the modules for planning and NLG, must
be tightly integrated.

From a planning perspective, the use of planning in GIVE imposes very
strict runtime requirements on the planner: Planning must happen in
real time, because if the system takes too much time between deciding
to give an instruction and actually giving this instruction, the user
may have walked or turned away, thus making the instruction invalid.
Furthermore, plan execution monitoring also plays an important role in
the GIVE problem.  At a high level, the system needs to monitor a
user's actions and compare them against the generated instruction set
to determine if the user has correctly followed directions or not. In
the case of the latter, new instructions may have to be generated. In
practice, the situation can be quite complicated since the mental
state of the user is not known and so the system must observe the
user's actions in real time. For example, a user directed to ``turn
around and walk through the door'' may not necessarily perform these
actions to the letter, i.e., immediately turning 180 degrees and
proceeding directly to the door. Instead, the user might take a
roundabout route through the room, eventually exiting out the
door. Although the user's actions do not match the generated
instructions exactly, they meet the intended goal. The system must be
able to identify such ``equivalent'' plans and not immediately
generate new instructions as soon as the user's actions have gone off
course. Furthermore, a user can communicate certain intentions to the
system, both through action and inaction. For instance, the system
should infer that a user has failed to follow instructions if the user
exits a room when given a directive to ``walk to the centre of the
room''. However, the system should also make a similar conclusion if a
user simply does nothing when given the instruction.

### 2008-06-23 trunk/Papers/experiences/experiments.tex
I'm essentially happy now

The results of this experiment are shown in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
(in lexicographic order) on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms
SGPLAN in this domain.\footnote{Experiments with SGPLAN use a
 pre-release version of SGPLAN, kindly provided by Chih-Wei Hsu. The
 release version of SGPLAN 5.2.2 had a bug, causing it to crash on some
 instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem to avoid unnecessary instantiations during search. In
particular, the ratio of grounding time to total runtime is generally above
85\%, and rises to above 99\% at $m=11$, which is still a small universe in
this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}  This is also reflected in the
much higher sensitivity of FF to the domain size: if we fix $n=1$, FF takes
60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$. By contrast,
our naive GraphPlan implementation takes 30 ms at $m=1$ and still only 50
ms at $m=10$.

Our results illustrate that FF is outperformed quite clearly by an ad-hoc
Java implementation of GraphPlan which only computes instances of
predicates and actions as they are discovered, while the planning graph is
being built. On the other hand, the runtime that FF actually spends on
search is again considerably lower than that of our GraphPlan
implementation, and for any given $m$, GraphPlan's runtime grows much
faster with $n$ (i.e., the plan size) than FF's. As a result, for
moderately-sized instances of the sentence planning domain FF does very
well on search, but spends about ten times as much runtime again on
grounding. Larger (but still realistically-sized) instances of the sentence
generation problem are still problematic for the planners we tested.

@@

In the second experiment, we evaluate the performance of the planners on
problems arising in the GIVE domain. We construct a series of test worlds,
similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
worlds consist of a $2n$ by $h$ grid of positions, such that there are
buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
player starts in position $(1,1)$ and must press all buttons in order to
successfully complete the game. The world is generated as a GIVE world
description, and then automatically converted into a planning problem by
the GIVE software. (We refer the reader to \todo{add citation} for more
details about this process.)

@@

Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$) are
shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
runs out of memory, this time at $w=17$. SGPLAN happily solves inputs
beyond $w=70$. However, unlike in Experiment 2, both planners now spend a
substantial proportion of their time on grounding. In SGPLAN, this
translates to a ``parsing time'' (which \todo{presumably} includes
grounding) which grows from 180 ms to 21.7 seconds as $w$ grows from $1$ to
$75$. The rest of the runtime (which also includes the search time) only
grows from 400 ms to 2.3 seconds. This difference is particularly dramatic
given that the actual optimal plan in each case is an identical plan of
about 100 steps. The planning times for these instances are also concerning
since times over a couple seconds negatively affect the overall response
time of the system, which must react in real time to user actions.

>>>>

The results of this experiment are shown in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
(in lexicographic order) on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms
SGPLAN in this domain.\footnote{Experiments with SGPLAN use a
 pre-release version of SGPLAN, kindly provided by Chih-Wei Hsu. The
 release version of SGPLAN 5.2.2 had a bug, causing it to crash on some
 instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem to avoid unnecessary instantiations during search. In
particular, the ratio of grounding time to total runtime is generally above
85\%, and rises to above 99\% at $m=11$, which is still a small universe in
this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}

In fact, FF spends so much time on grounding that it is consistently
outperformed by an ad-hoc Java implementation of GraphPlan which only
computes instances of predicates and actions as they are discovered,
while the planning graph is being built -- despite the fact that FF is
consistently much faster as far as pure search time is concerned.
Looking at the difference from a different angle, FF's performance is
much more sensitive to the domain size: if we fix $n=1$, FF takes 60
ms to compute a plan at $m=1$, but 2.4 seconds (for the same plan) at
$m=10$; our GraphPlan implementation takes 30 ms at $m=1$ and still
only 50 ms at $m=10$.  Conversely, GraphPlan's runtime grows much
faster with the plan size (i.e., with growing values of $n$ for a
fixed $m$).  Larger (but still realistically-sized) instances of the
sentence generation problem are still problematic for the planners we
tested.

@@

In the second experiment, we evaluate the performance of the planners
on problems arising in the GIVE domain. We construct a series of test
worlds, similar to the one illustrated in
Figure~\ref{fig:give-minimal}. These worlds consist of a $2n$ by $h$
grid of positions, such that there are buttons at positions $(2i-1,1)$
and $(2i,h)$ for $1 \leq i \leq n$. The player starts in position
$(1,1)$ and must press all buttons in order to successfully complete
the game. The world is generated as a GIVE world description, and then
automatically converted into a planning problem as in
Fig.~\ref{fig:give-planning} by the GIVE software.

@@

Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$)
are shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2,
FF again runs out of memory, this time at $w=17$. SGPLAN happily
solves inputs beyond $w=70$. However, unlike in Experiment 2, both
planners now spend a substantial proportion of their time on
grounding. In SGPLAN, this translates to a ``parsing time'' (which we
assume includes grounding) which grows from 180 ms to 21.7 seconds as
$w$ grows from $1$ to $75$. The rest of the runtime (which also
includes the search time) only grows from 400 ms to 2.3 seconds. This
difference is particularly dramatic given that the actual optimal plan
in each case is an identical plan of about 100 steps. The planning
times for these instances are also concerning since times over a
couple seconds negatively affect the overall response time of the
system, which must react in real time to user actions.

### 2008-06-23 trunk/Papers/experiences/introduction.tex
small change
I'm essentially happy now

Natural language generation (NLG; \citeauthor{reiter00building}
\citeyear{reiter00building}) is one of the major subfields of natural
language processing, concerned with computing natural language sentences or
texts that convey a given piece of information to the user. Intuitively,
this task can be viewed as a problem involving actions, beliefs, and goals:
an agent communicating with another agent tries to change the mental state
of the hearer by applying actions which correspond to the utterance of
words or sentences. This characterization of NLG suggests obvious parallels
to automated planning---a view which has a long tradition in NLG
\cite{appelt:planning,young94dpocl}.  More recently, there has been a
resurgence of interest in applying modern planning techniques to NLG
\cite{Steedman-Petrick:07,KolSto07,benotti08b}.  While efficiency has not
been the main focus of NLG, problems in NLG are often more complex than
their counterparts in other areas of natural language processing (such as
parsing \cite{KolStr02}), giving rise to computationally challenging
domains. These new approaches combine an interest in planning as a
modelling tool for natural language, with a hope of improved efficiency by
exploiting modern approaches and algorithms.

@@

We will then discuss some of our experiences using off-the-shelf planners
in these domains. In particular, we explore the efficiency of FF
\cite{HoffmannNebel01} and SGPLAN
\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
on traditional benchmarks, and in the International Planning Competition
(IPC)---on a range of problem instances from our planning domains.  We also
compare these results against an ad-hoc Java implementation of GraphPlan.
Our findings are mixed. On the one hand, it turns out that modern planners
handle some of the search problems that arise in NLG quite easily (although
large problem instances remain a challenge). On the other hand, these same
planners spend tremendous amounts of time on preprocessing. For instance,
FF spends 90\% of its runtime in the sentence generation domain on
grounding out literals and actions, and ends up slower than our version of
GraphPlan that avoids grounding. For domains like ours, which are dominated
by the number of actions and the universe size, rather than the
combinatorics of the search problem, this observation suggests that the
overall runtime of a planner could be improved by grounding more
selectively. We therefore offer these domains as challenges for the
planning community.

>>>>

Natural language generation (NLG; \citeauthor{reiter00building}
\citeyear{reiter00building}) is one of the major subfields of natural
language processing, concerned with computing natural language
sentences or texts that convey a given piece of information to the
user. Intuitively, this task can be viewed as a problem involving
actions, beliefs, and goals: an agent communicating with another agent
tries to change the mental state of the hearer by applying actions
which correspond to the utterance of words or sentences. This
characterization of NLG suggests obvious parallels to automated
planning---a view which has a long tradition in NLG
\cite{appelt:planning,young94dpocl}.  More recently, there has been a
resurgence of interest in applying modern planning techniques to NLG
\cite{Steedman-Petrick:07,KolSto07,benotti08b}.  While efficiency has
not always been the main focus of NLG, problems in NLG are often more
complex than their counterparts in other areas of natural language
processing (such as parsing \cite{KolStr02}), giving rise to
computationally challenging domains. These new approaches combine an
interest in planning as a modelling tool for natural language, with a
hope of improved efficiency by exploiting modern approaches and
algorithms.

@@

We will then discuss some of our experiences using off-the-shelf planners
in these domains. In particular, we explore the efficiency of FF
\cite{HoffmannNebel01} and SGPLAN
\cite{hsu06:_new_featur_in_sgplan_for}---planners that have been successful
on traditional benchmarks, and in the International Planning Competition
(IPC)---on a range of problem instances from our planning domains.  We also
compare these results against an ad-hoc Java implementation of GraphPlan
\cite{Blum1997}.  Our findings are mixed. On the one hand, it turns out
that modern planners handle some of the search problems that arise in NLG
quite easily (although large problem instances remain a challenge). On the
other hand, these same planners spend tremendous amounts of time on
preprocessing. For instance, FF spends 90\% of its runtime in the sentence
generation domain on grounding out literals and actions, and ends up slower
than our version of GraphPlan that avoids grounding. For domains like ours,
which are dominated by the number of actions and the universe size, rather
than the combinatorics of the search problem, this observation suggests
that the overall runtime of a planner could be improved by grounding more
selectively. We therefore offer these domains as challenges for the
planning community.
